### 2014-12-14 text/Chapter06.tex
Finishing proofreading (Chapter 6)

\chapter{Unified coeffect language ???} 
\label{ch:unified}

@@

\section{Introduction}

The main goal of this thesis is to provide a \emph{unified} calculus for tracking context 
dependence. We have not achieved this goal yet. In Chapter~\ref{ch:applications}, we identified 
two kinds of contextual properties that we further covered separately -- flat coeffects in 
Chapter~\ref{ch:flat} track whole-context properties and structural coeffects, covered in 
Chapter~\ref{ch:structural}, track per-variable properties. In this chapter, we unify
the two notions. We introduce a \emph{unified coeffect} calculus that generalizes the two systems
and can be instantiated to track both flat and structural properties (Section~\ref{sec:unified-unified}).

Although the results presented in this thesis are mainly of a theoretical nature, we indeed 
believe that coeffects should be integrated in main-stream programming languages. In the second
part of this chapter, we discuss an alternative approach to defining coeffect 
systems which highlights the relationship between our work and related work arising from modal logics 
(Section~\ref{sec:unified-meta}). Finally, we outline one possible approach for practical
implementations of coeffects (Section~\ref{sec:unified-impl}).

@@

The flat coeffect calculus (Figure~\ref{fig:flat-types}) and the structural coeffect calculus 
differ in a number of ways (Figure~\ref{fig:struct-types}). Understanding the differences is the 
key to reconciling the two systems:

\begin{itemize}
\item Structural coeffect calculus contains explicit rules for context manipulation
  (weakening, contraction, exchange). In the flat coeffect calculus, these rules are not defined
  explicitly, but are admissible.

\item In the structural coeffect calculus, the variable context is treated as a vector
  and is annotated with a vector of (scalar) coeffects. In the flat coeffect calculus,
  the variable context is a set and is annotated with a single (scalar) coeffect.

\item In the flat coeffect calculus, we distinguish between splitting of the context requirements
  and merging of context requirements ($\cpar$ and $\czip$, respectively). In the structural
  coeffect calculus, the operations (which model splitting and appending vectors) are invertible 
  and so the structural coeffect algebra requires just $\atimes$.
\end{itemize}

\noindent
In the unified calculus presented in this section, we address the three differences as follows.
We use calculus with explicit rules for context manipulation. In systems that arise from flat
calculi, the rules can be applied freely without changing the coeffects. We generalize the 
structure of coeffect annotations using the notion of a ``container'' which can be specialized 
to obtain a single annotation or a vector of annotations. Finally, we distinguish between 
splitting and merging of context requirements (using the notation $\apar$ and $\azip$, 
respectively). For structural coeffect calculi, the two operators coincide, but for flat
coeffect calculi, they provide the needed flexibility.

\subsection{Shapes and containers}

Our notion of a \emph{coeffect container} is based on the idea of a container introduced by 
Abbott et al. \cite{types-containers}. Interestingly, the work on containers has later been linked
to comonads by Ahman et al. \cite{comonads-containers}. Intuitively a container describes data
types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of lists).
For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
specified length). More formally:

\begin{definition}
A \emph{container} $S \triangleleft P$ is given by a set $S$ of shapes and a shape-indexed family
$P : S \rightarrow \ident{Set}$ of positions.
\end{definition}

\noindent
Well-known examples of containers include lists, non-empty lists, (un\-boun\-ded) streams and 
singleton data type (which contains exactly one element). Two containers relevant to our work
are lists and singleton data types:

\begin{itemize}
\item The container representing lists is given by $S \triangleleft P$ where 
  shapes are integers $S = \ident{Nat}$ (lengths of a list). The set of positions for a 
  given length $n$ is a set of indices $P n = \{ 1 \ldots n \}$.

\item The contianer representing singleton data type is given by $S \triangleleft P$ where
  shapes are given by a singleton set $S = \{ \ast \}$ and the set of positions for the
  shape $\ast$ contains exactly one position $P \ast = \{ 0 \}$.
\end{itemize}

@@

\subsection{Structure of coeffects}
In the structural coeffect calculus, the structure annotation was formed by a vector of coeffect
sclars. The unified coeffect calculus is similar, but a \emph{vector} is replaced with a 
\emph{container}. The primitive coeffect annotations in the unified calculus are formed by
a \emph{coeffect scalar}, which remains the same as in structural coeffect calculus
(Definition~\ref{def:structural-scalar}). In this section, we refer to it as
\emph{\cclrd{unified coeffect scalar}} (and we repeat the definition below). Then we
define \emph{\sclrd{unified coeffect containers}} which determines how coeffect scalar values 
are attached to the free-variable context. Finally, we define the \emph{\aclrd{unified coeffect 
algebra}} which consists of shape-indexed coeffect scalar values.

@@

\paragraph{Coeffect scalar.}
The following definition of the coeffect scalar structure repeats the Definition~\ref{def:structural-scalar}
from the previous chapter.

\begin{definition}
A \emph{\cclrd{unified coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set 
$\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations 
$\cseq, \cpar$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and
$(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
%
\begin{equation*}
\begin{array}{ccr}
r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
\textnormal{(monoid)}   
\\
r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
\czero \;\cpar\; r = r = r \;\cpar\; \czero &
\textnormal{(monoid)}   
\\
\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
t\; \cleq\; t &
\textnormal{(pre-order)}   
\end{array}
\end{equation*}
\end{definition}

@@

The $\cpar$ operation models combining of context requirements arising from multiple parts of a 
program. The meaning depends on the coeffect container. The operation can either combine requirements
of individual variables (structural) or requirements attached to the whole context of multiple
sub-exp\-ressions (flat).

\paragraph{Coeffect containers.}
The coeffect container is a container that determines how are scalar coeffect annotations attached 
to free-variable contexts. In addition to a container $\SHP$ with shapes and shape-indexed positions, the
coeffect container also provides a mapping that returns the shape of a free-variable context. The
mapping between the shape of the variable context and the shape of the coeffect annotation
is not necessarily bijective. For example, coeffect annotations in flat systems have just a single 
shape $\SH = \{ \ast \}$.

In the coeffect judgment $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$, the coeffect annotation
$\aclrd{\textbf{r}}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape of $\Gamma$.
We write $\sclrd{s} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$. The operation $\SP\sclrd{s}$
returns a \emph{set} of positions and so we can write $\aclrd{\textbf{r}} \in \SP\sclrd{s} \rightarrow \C$ 
as a mapping from positions (defined by the shape) to scalar coeffects. We usually write this as 
the exponent $\aclrd{\textbf{r}} \in \C^{\SP\sclrd{s}}$.

The coeffect container is also equipped with an operation that appends shapes (when we concatenate
variable contexts) and two special shapes in $\SH$ representing empty context and singleton context.

@@

The elements $\sempty$ and $\sunit$ represent the shapes of empty and singleton free-variable contexts 
respectively. The $\stimes$ operation corresponds to concatentation of free-variable contexts. Given
$\Gamma_1$ and $\Gamma_2$ such that $\sclrd{s_1}=\slen{\Gamma_1}, \sclrd{s_2}=\slen{\Gamma_2}$, 
we require that $\sclrd{s_1}\,\stimes\,\sclrd{s_2}=\slen{\Gamma_1, \Gamma_2}$. 
\end{definition}

\noindent
As said earlier, we use two kinds of coeffect containers that describe the structure of vectors 
(for structural coeffects) and the shape of trivial singleton container (for flat coeffects):

\begin{example}
\label{ex:unified-struct-shape}
Structural coeffect container is defined as $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$
where $\SH=\mathbb{N}$ and $\SP \sclrd{n}=\{ 1 \ldots n \}$. The shape mapping ${|}\Gamma{|}$ returns the 
number of variables in $\Gamma$. Empty and singleton contexts are annotated with $0$ and $1$, 
respectively, and shapes of combined contexts are added so that ${|}\Gamma_1, \Gamma_2{|} = 
{|}\Gamma_1{|} + {|}\Gamma_2{|}$.

Therefore, a coeffect annotation is a \emph{vector} 
$\aclrd{\textbf{r}} \in \C^{\SP\sclrd{n}}$ and assigns a coeffect scalar $\aclrd{\textbf{r}}(i) \in \C$ 
for each position (corresponding to a variable $x_i$ in the context).
\end{example}

\begin{example}
\label{ex:unified-flat-shape}
Flat coeffect container is defined as $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$.
The container is defined as a singleton data type $\SH = \{\sflat \}$ and $\SP \sflat = \{ 0 \}$
with a constant function ${|}\Gamma{|}=\sflat$ and a trivial operation $\sflat\,\diamond\,\sflat = \sflat$.

That is, there is a single shape
$\sflat$ with a single position and all free-variable contexts have the same singleton shape.
Therefore, a coeffect annotation is drawn from $\C^\{\sflat\}$ which is isomorphic to $\C$
and so a coeffect scalar $\cclrd{r}\in \C$ is associated with every free-variable context.
\end{example}

\newpage
\begin{example}
Similarly to the previous example, we can also define a coeffect container with \emph{no} positions,
\ie~$\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ where $\SH = \{\sflat \}$, $\SP \sflat = \emptyset$,
${|}\Gamma{|}=\sflat$ and $\sflat\,\diamond\,\sflat = \sflat$.

This reduces our system to the simply-typed $\lambda$-calculus with no context annotations, because
$\SP\sflat=\emptyset$ and so coeffect annotations would be from the set $\C^\emptyset$.
\end{example}

\paragraph{Unified coeffect algebra.}
The coeffect calculus annotates judgments with shape-indexed coeffect annotations.
The \emph{\aclrd{unified coeffect algebra}} combines a coeffect scalar and coeffect container
to define shape-indexed coeffects and operations for manipulating these.

The definition here reconciles the third point discussed in Section~\ref{sec:unified-unified} --
the fact that flat coeffects use separate operations for splitting and merging ($\cpar$ and $\czip$)
while structural coeffects use tensor $\atimes$. In the unified calculus, we use two operators
that can, however, be coincide.

\begin{definition}
Given a \cclrd{unified coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a 
\sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \emph{\aclrd{unified 
coeffect algebra}} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP \sempty}$
is a coeffect annotation for the empty context and $\azip, \apar$ are families of operations that 
combine coeffect annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m} \in \SH$:
%
\begin{equation*}
\apar_{\sclrd{m},\sclrd{n}}, \azip_{\sclrd{m},\sclrd{n}} ~:~ \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})}
\end{equation*}
\end{definition}

\noindent
A coeffect algebra induces the following three additional operations:
%
\begin{equation*}
\begin{array}{rclcrcl}
 \alift{-} &\narrow{:}& \C \rightarrow \C^{\SP \sunit}  \\
 \alift{\cclrd{x}} &\narrow{=}& \lambda \_ . \cclrd{x}   \\
\\[-0.5em]
 \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\SP \sclrd{m}} \rightarrow \C^{\SP \sclrd{m}}   \\
 \cclrd{r}\, \aseq\, \aclrd{\textbf{s}} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\, \cseq\, \aclrd{\textbf{s}}(\sclrd{s}) \\
\\[-0.5em]
 \alen{-} &\narrow{:}& \C^{\SP \sclrd{m}} \rightarrow \sclrd{m}\\
 \alen{\aclrd{\textbf{r}}} &\narrow{=}& \sclrd{m}
\end{array}
\end{equation*}
%
The $\alift{-}$ operation lifts a scalar coeffect to a shape-indexed coeffect that is indexed by 
the singleton context shape. The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a 
vector by scalar. As we always use bold font for vectors and ordinary font for scalars (as well 
as distinct colour), using the same symbol is not ambiguous. We also tend to omit the subscript
$\sclrd{m}$ and write $\aseq$.

Finally, we define $\alen{}$ as an operation that returns the shape of a given shape-indexed 
coeffect. The only purpose is to simplify notation, as we often need to specify that shapes of 
variable context and coeffect match, \eg~$\alen{\aclrd{\textbf{r}}} = \slen{\Gamma}$.

\paragraph{Splitting and merging coeffects.}
The operators $\apar$ and $\azip$ combine shape-indexed coeffects associated with two contexts. 
For example, assume we have $\Gamma_1$ and $\Gamma_2$ with coeffects $\aclrd{\textbf{r}} \in 
\C^{\SP\sclrd{m}}$ and $\aclrd{\textbf{s}} \in \C^{\SP\sclrd{n}}$. In the structural system, the 
context shapes $\sclrd{m}, \sclrd{n}$ denote the number of variables in the two contexts. The 
combined context $\Gamma_1, \Gamma_2$ has a shape $\sclrd{m}\,\stimes\,\sclrd{n}$ and the combined 
coeffects $\aclrd{\textbf{r}} \,\azip\, \aclrd{\textbf{s}}, \aclrd{\textbf{r}} \,\apar\, \aclrd{\textbf{s}} 
\in \C^{\SP(\sclrd{m} \stimes \sclrd{n})}$ are indexed by that shape.

@@

\vspace{0.5em}
\caption{Type system for the unified coeffect calculus}
\label{fig:unif-types}
\end{figure}

@@

\subsection{Unified coeffect type system}
\label{sec:unified-types}
The unified coeffect system in Figure~\ref{fig:unif-types} resembles the structural type system
shown in Figure~\ref{fig:struct-types}. Rather than explaining the rules one-by-one, we focus on
the key differences between the two.

The type system for the unified coeffect calculus is parameterized by a coeffect scalar 
$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ together with a coeffect algebra $(\azip, \apar, \azero)$ 
and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$. 
As in the structural system, free-variable contexts $\Gamma$ are treated as vectors modulo duplicate
use of variables -- the associativity is built-int. The order of variables matters, but can be changed 
using the structural rules. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}} \in \C^\SH$ 
are shape-indexed coeffects (rather than simple vectors as before). As before, functions are
annotated with coeffects scalars. % $\cclrd{r}, \cclrd{s}, \cclrd{t} \in \C$.

\paragraph{Syntax-driven rules.} 
The (\emph{var}) rule is syntactically the same as in the structural system, but it should be read
differently. The $\alift{-}$ operation does not create a \emph{vector}, but a \emph{shape-indexed
coeffect} that returns $\cunit$ for all positions in the singleton shape $\sunit$. 
The (\emph{const}) rule annotates empty context with a special annotation of the unit shape.

@@

In the (\emph{app}), (\emph{abs}) and (\emph{let}) rules, the only change from the structural
system is that vector concatentation $\atimes$ is now replaced with context splitting/merging 
of the unified coeffect algebra. In particular, we use splitting of context requirements $\azip$
in rule \emph{assumptions} and merging of context requirements $\apar$ in rule \emph{conclusions}.

@@

The unified coeffect system uses a general notion of context shape, but it has been designed with 
structural and flat systems in mind. In this and the next section, we show how it captures 
the two kinds of systems.

>>>>

\chapter{Unified coeffect languages} 
\label{ch:unified}

@@

The main goal of this thesis is to provide a \emph{unified} calculus for tracking context 
dependence. We have not achieved this goal yet. In Chapter~\ref{ch:applications}, we identified 
two kinds of contextual properties that we further covered separately. Flat coeffects in 
Chapter~\ref{ch:flat} track whole-context properties, while structural coeffects, covered in 
Chapter~\ref{ch:structural}, track per-variable properties. In this chapter, we unify the two 
notions.

Although the results presented in this thesis are mainly of a theoretical nature, we indeed 
believe that coeffects should be integrated in main-stream programming languages. This is addressed
in the second part of the chapter, where we outline possible approach for a practical implementation
of coeffect systems. Finally, we conclude the chapter with a brief discussion of an alternative
approach to coeffect system, based on the meta-language style.

\paragraph{Chapter structure and contributions}
\begin{itemize}
\item We start by unifying the two kinds of coeffects discussed so far. In 
  Section~\ref{sec:unified-unified}, we introduce the \emph{unified coeffect calculus}, which 
  generalizes flat and structural systems and can be instantiated to track both flat and structural 
  properties.

\item We outline one possible approach for practical implementations of coeffects 
  (Section~\ref{sec:unified-impl}). The technique is based on embedding comonadic computations
  via a lightweight syntactic extension and using rich type systems for capturing the structure of 
  coeffect algebra.

\item Finally, we discuss an alternative approach to defining coeffect systems based on the
  meta-language style, which highlights the relationship between our work and related work arising 
  from modal logics such as CMTT \cite{logic-cmtt} (Section~\ref{sec:unified-meta}).
\end{itemize}

\noindent
The unified coeffect calculus presented in the first part of the chapter is a novel result that
completes the theory developed in this thesis. The next two sections present important future and
related work, respectively.

@@

The type systems of the flat coeffect calculus (Figure~\ref{fig:flat-types}) and the structural 
coeffect calculus (Figure~\ref{fig:struct-types}) differ in a number of ways. Understanding the 
differences is the key to reconciling the two systems:

\begin{itemize}
\item The structural coeffect calculus contains explicit rules for context manipulation
  (weakening, contraction, exchange). In the flat coeffect calculus, these rules are not defined
  explicitly, but are admissible.

\item In the structural coeffect calculus, the variable context is treated as a vector
  and is annotated with a vector of (scalar) coeffects. In the flat coeffect calculus,
  the variable context is a finite partial function (from names to types) and is annotated with 
  a single (scalar) coeffect.

\item In the flat coeffect calculus, we distinguish between \emph{splitting} context requirements
  and \emph{merging} of context requirements ($\cpar$ and $\czip$, respectively). In the structural
  coeffect calculus, the operations (which model appending and splitting vectors) are invertible 
  and so the structural coeffect algebra uses the same tensor product $\atimes$.
\end{itemize}

\noindent
In the unified calculus presented in this section, we address the three differences as follows.
We include explicit rules for context manipulation in the calculus; in systems that arise from flat
calculi, the structural rules do not change coeffects and can thus be applied freely. We generalize 
the structure of coeffect annotations using the notion of a ``container'' which can be specialized 
to obtain a single annotation or a vector of annotations. This could potentially be used to capture
other structures such as trees in bunched typing \cite{substruct-bunched}. Finally, we distinguish 
between splitting and merging of context requirements (using the notation $\apar$ and $\azip$, 
respectively). For structural coeffect calculi, the two operators coincide, but for flat
coeffect calculi, they provide the needed flexibility.

\subsection{Shapes and containers}
Our generalization of coeffect structure using a \emph{coeffect container} is based on containers of 
Abbott et al. \cite{types-containers}. Interestingly, containers have later been linked
to comonads by Ahman et al. \cite{comonads-containers}. Intuitively a container describes data
types such as lists, trees or streams. A container is formed by shapes (\eg~lengths of 
lists)\footnote{Shapes can also be intuitively thought of as sizes, but this is not fully precise 
as some containers may have multiple shapes of the same length, \eg~differently ballanced trees.}.
For every shape, we can obtain a set of positions in the container (\eg~offsets in a list of a
specified length). More formally:

\begin{definition}
A \emph{container} consists of a pair $(S,P)$, usually written as $S \triangleleft P$. 
Here, $S$ is a set of shapes and $P$ is a shape-indexed family $P : S \rightarrow \ident{Set}$ 
of positions.
\end{definition}

\noindent
Well-known examples of containers include lists, non-empty lists, (un\-boun\-ded) streams,
trees and the singleton data type (which contains exactly one element). Containers relevant 
to our work are lists and singleton data types:

\begin{itemize}
\item The container representing lists is given by $S \triangleleft P$ where 
  shapes are integers $S = \ident{Nat}$ (lengths of a list). The set of positions for a 
  given length $n$ is the set of indices $P(n)= \{ 1 \ldots n \}$.

\item The contianer representing the singleton data type is given by $S \triangleleft P$ where
  shapes are given by a singleton set $S = \{ \ast \}$ and the set of positions for the
  shape $\ast$ contains exactly one position. To follow the intuition based on offset or 
  index, we write the single position as $0$, that is $P(\ast) = \{ 0 \}$.
\end{itemize}

@@

\subsection{Structure of coeffects}
In the structural coeffect calculus, the coeffect annotation was formed by a vector of coeffect
scalars. The annotations in the unified coeffect calculus are similar, but a \emph{vector} is 
replaced with a more general \emph{container}. The primitive coeffect annotations in the unified 
calculus are formed of \emph{coeffect scalars}. The coeffect scalars are equipped with the same
structure as in structural coeffect calculus (Definition~\ref{def:structural-scalar}). In this 
section, we refer to it as \emph{\cclrd{unified coeffect scalar}} (and we repeat the definition 
below). Then we define \emph{\sclrd{unified coeffect containers}} which determines how coeffect 
scalar values are attached to the free-variable context. Finally, we define the 
\emph{\aclrd{unified coeffect algebra}} which consists of shape-indexed coeffect scalar values.

@@

\paragraph{Coeffect scalars.}
The following definition of the coeffect scalar structure repeats the 
Definition~\ref{def:structural-scalar} from the previous chapter.

\begin{definition}
A \emph{\cclrd{unified coeffect scalar}} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ is a set 
$\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations 
$\cseq, \cpar$ such that $(\C, \cseq, \cunit)$ and $(\C, \cpar, \czero)$ are monoids and
$(\C, \cleq)$ is a pre-order. That is, for all $r,s,t\in \C$:
%
\begin{equation*}
\begin{array}{ccr}
r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t  &
\cunit \;\cseq\; r = r = r \;\cseq\; \cunit &
\textnormal{(monoid)}   
\\
r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
\czero \;\cpar\; r = r = r \;\cpar\; \czero &
\textnormal{(monoid)}   
\\
\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t&
t\; \cleq\; t &
\textnormal{(pre-order)}   
\\
\end{array}
\end{equation*}
%
The following distributivity axioms hold:
\begin{align*}
\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
\end{align*}
%
In addition, we require the following two properties of $\czero, \cpar$ and $\cleq$:
\begin{equation*}
\begin{array}{l}
 \quad \cclrd{r}\,\cleq\,\cclrd{r'} \; \Rightarrow \; \forall\cclrd{s}.\;(\cclrd{r}\,\cseq\,\cclrd{s})\;\cleq\;(\cclrd{r'}\,\cseq\,\cclrd{s}) \\
 \quad \czero\;\cleq\;(\czero\,\cseq\,\cclrd{r})
\end{array}
\end{equation*}
\end{definition}

@@

The $\cpar$ operation models combining of context requirements arising from multiple parts of a 
program. The meaning depends on the coeffect container. The operation can either combine requirements
of individual variables (structural coeffects) or requirements attached to the whole context of 
multiple sub-exp\-ressions (flat coeffects).

\paragraph{Coeffect containers.}
The coeffect container is a container that determines how are scalar coeffects attached to 
free-variable contexts. In addition to a container $\SHP$ formed by shapes and shape-indexed positions, 
the coeffect container provides a mapping that returns the shape corresponding to a free-variable context.

The mapping between the shape of the variable context and the shape of the coeffect annotation
is not necessarily bijective. For example, coeffect annotations in flat systems have just a single 
shape $\SH = \{ \ast \}$.

In the coeffect judgment $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$, the coeffect annotation
$\aclrd{\textbf{r}}$ is drawn from the set of coeffect scalars $\C$ indexed by the shape corresponding
to $\Gamma$. We write $\sclrd{s} = \slen{\Gamma}$ for the shape corresponding to $\Gamma$. The operation 
$\SP(\sclrd{s})$ returns a \emph{set} of positions and so we can write $\aclrd{\textbf{r}} \in 
  \SP(\sclrd{s}) \rightarrow \C$  as a mapping from positions (defined by the shape) to scalar coeffects. 
We usually write this as the exponent $\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{s})}$.
The coeffect container is also equipped with an operation that appends shapes (when we concatenate
variable contexts) and two special shapes in $\SH$ representing the empty context and the singleton 
context.

@@

The elements $\sempty$ and $\sunit$ represent the shapes of the empty and the singleton free-variable 
context, respectively. The $\stimes$ operation corresponds to concatentation of free-variable contexts.
Given $\Gamma_1$ and $\Gamma_2$ such that $\sclrd{s_1}=\slen{\Gamma_1}, \sclrd{s_2}=\slen{\Gamma_2}$, 
we require that $\sclrd{s_1}\,\stimes\,\sclrd{s_2}=\slen{\Gamma_1, \Gamma_2}$. 
\end{definition}

\noindent
As discussed earlier, we use two kinds of coeffect containers that describe the structure of vectors 
(for structural coeffects) and the shape of trivial singleton container (for flat coeffects):

\begin{example}
\label{ex:unified-struct-shape}
Structural coeffect container is defined as $\sclrd{(\SHP, {|}\textnormal{--}{|}, +, 0, 1)}$
where $\SH=\mathbb{N}$ and $\SP(\sclrd{n})=\{ 1 \ldots n \}$. The shape mapping ${|}\Gamma{|}$ returns the 
number of variables in $\Gamma$. Empty and singleton contexts are annotated with $0$ and $1$, 
respectively, and shapes of combined contexts are added so that ${|}\Gamma_1, \Gamma_2{|} = 
{|}\Gamma_1{|} + {|}\Gamma_2{|}$.

Therefore, a coeffect annotation is a \emph{vector} 
$\aclrd{\textbf{r}} \in \C^{\SP(\sclrd{n})}$ and assigns a coeffect scalar $\aclrd{\textbf{r}}(i) \in \C$ 
for each position (corresponding to a variable $x_i$ in the context).
\end{example}

\begin{example}
\label{ex:unified-flat-shape}
Flat coeffect container is defined as $\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$.
The container is defined as a singleton data type $\SH = \{\sflat \}$ and $\SP(\sflat) = \{ 0 \}$
with a constant function ${|}\Gamma{|}=\sflat$ and a trivial operation $\sflat\,\diamond\,\sflat = \sflat$.

That is, there is a single shape
$\sflat$ with a single position and all free-variable contexts have the same singleton shape.
Therefore, a coeffect annotation is drawn from $\C^{\{\sflat\}}$ which is isomorphic to $\C$
and so a coeffect scalar $\cclrd{r}\in \C$ is associated with every free-variable context.
\end{example}

\begin{example}
Similarly to the previous example, we can also define a coeffect container with \emph{no} positions,
\ie~$\sclrd{(\SHP, {|}-{|}, \diamond, \sflat, \sflat)}$ where $\SH = \{\sflat \}$, $\SP(\sflat) = \emptyset$,
${|}\Gamma{|}=\sflat$ and $\sflat\,\diamond\,\sflat = \sflat$. 
This reduces our system to the simply-typed $\lambda$-calculus with no context annotations, because
$\SP(\sflat)=\emptyset$ and so coeffect annotations would be from the empty set $\C^\emptyset$.
\end{example}

\paragraph{Unified coeffect algebra.}
The unified coeffect calculus annotates typing judgments with shape-indexed coeffect annotations.
The \emph{\aclrd{unified coeffect algebra}} combines a coeffect scalar and a coeffect container
to define shape-indexed coeffects and operations for manipulating these.

The definition here reconciles the third point discussed in Section~\ref{sec:unified-unified} --
the fact that flat coeffects use separate operations for splitting and merging ($\cpar$ and $\czip$)
while structural coeffects use the tensor $\atimes$. In the unified calculus, we use two operators
that can, however, coincide.

\begin{definition}
Given a \cclrd{unified coeffect scalar} $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and a 
\sclrd{coeffect container} $(\SHP, \slen{-}, \stimes, \sempty, \sunit)$ a \emph{\aclrd{unified 
coeffect algebra}} extends the two structures with $(\azip, \apar, \azero)$ where $\azero \in \C^{\SP(\sempty)}$
is a coeffect annotation for the empty context and $\azip, \apar$ are families of operations that 
combine coeffect annotations indexed by shapes. That is $\forall \sclrd{n},\sclrd{m} \in \SH$:
%
\begin{equation*}
\apar_{\sclrd{m},\sclrd{n}}, \azip_{\sclrd{m},\sclrd{n}} ~:~ \C^{\SP \sclrd{m}} \times \C^{\SP \sclrd{n}} \rightarrow \C^{\SP (\sclrd{m}\,\stimes\,\sclrd{n})}
\end{equation*}
\end{definition}

\noindent
A coeffect algebra induces the following three additional operations:
%
\begin{equation*}
\begin{array}{rclcrcl}
 \alift{-} &\narrow{:}& \C \rightarrow \C^{\SP(\sunit)}  \\
 \alift{\cclrd{x}} &\narrow{=}& \lambda \_ . \cclrd{x}   \\
\\[-0.5em]
 \aseq_{\sclrd{m}} &\narrow{:}&  \C \times \C^{\SP(\sclrd{m})} \rightarrow \C^{\SP(\sclrd{m})}   \\
 \cclrd{r}\, \aseq\, \aclrd{\textbf{s}} &\narrow{=}& \lambda \sclrd{s} . \cclrd{r}\, \cseq\, \aclrd{\textbf{s}}(\sclrd{s}) \\
\\[-0.5em]
 \alen{-} &\narrow{:}& \C^{\SP(\sclrd{m})} \rightarrow \sclrd{m}\\
 \alen{\aclrd{\textbf{r}}} &\narrow{=}& \sclrd{m}
\end{array}
\end{equation*}
%
The $\alift{-}$ operation lifts a scalar coeffect to a shape-indexed coeffect that is indexed by 
the shape of a singleton context. The $\aseq_{\sclrd{m}}$ operation is a left multiplication of a 
vector by a scalar. As we always use bold face for vectors and ordinary face for scalars (as well 
as a distinct colour), using the same symbol is not ambiguous. We also tend to omit the subscript
$\sclrd{m}$ and write~$\aseq$.

Finally, we define $\alen{-}$ as an operation that returns the shape of a given shape-indexed 
coeffect. The only purpose is to simplify notation, as we tend to avoid subscripts, but often need 
to specify that shapes of variable context and coeffect match, 
\eg~$\alen{\aclrd{\textbf{r}}} = \slen{\Gamma}$.

\paragraph{Splitting and merging coeffects.}
The operators $\apar$ and $\azip$ combine shape-indexed coeffects associated with two contexts. 
For example, assume we have $\Gamma_1$ and $\Gamma_2$ with coeffects $\aclrd{\textbf{r}} \in 
\C^{\SP(\sclrd{m})}$ and $\aclrd{\textbf{s}} \in \C^{\SP(\sclrd{n})}$. In the structural system, the 
context shapes $\sclrd{m}, \sclrd{n}$ denote the number of variables in the two contexts. The 
combined context $\Gamma_1, \Gamma_2$ has a shape $\sclrd{m}\,\stimes\,\sclrd{n}$ and the combined 
coeffects $\aclrd{\textbf{r}} \,\azip\, \aclrd{\textbf{s}}, \aclrd{\textbf{r}} \,\apar\, \aclrd{\textbf{s}} 
\in \C^{\SP(\sclrd{m} \stimes \sclrd{n})}$ are indexed by that shape.

@@

\vspace{0.5em}
\figcaption{Type system for the unified coeffect calculus}
\label{fig:unif-types}
\end{figure}

@@

\subsection{Unified coeffect type system}
\label{sec:unified-types}
The unified coeffect system in Figure~\ref{fig:unif-types} resembles the structural type system
shown in Figure~\ref{fig:struct-types}. Rather than explaining the rules one-by-one, we focus on
how the unified system differs from the structural system.

The type system for the unified coeffect calculus is parameterized by a coeffect scalar 
$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ together with a coeffect algebra $(\azip, \apar, \azero)$ 
and the derived constructs $\alift{-}$, $\alen{-}$ and $\aseq$. 
As in the structural system, free-variable contexts $\Gamma$ are treated as vectors modulo duplicate
use of variables -- the associativity is built-in. The order of variables matters, but can be changed 
using the exchane rule. The context annotations $\aclrd{\textbf{r}}, \aclrd{\textbf{s}}, \aclrd{\textbf{t}}$ 
are shape-indexed coeffects (rather than simple vectors as before). As before, functions are
annotated with coeffects scalars.

\paragraph{Syntax-driven rules.} 
The (\emph{var}) rule is syntactically the same as in the structural system, but it should be read
differently. The $\alift{-}$ operation does not create a \emph{vector}, but a \emph{shape-indexed
coeffect} that returns $\cunit$ for all positions in the singleton shape $\sunit$. 
The (\emph{const}) rule annotates empty context with a special annotation of the shape $\sempty$.

@@

In the (\emph{app}), (\emph{abs}) and (\emph{let}) rules, the only change from the structural
system is that the vector concatentation $\atimes$ is now replaced with context splitting/merging 
of the unified coeffect algebra. As already mentioned, we use splitting of context requirements $\azip$
in rule \emph{assumptions} and merging of context requirements $\apar$ in rule \emph{conclusions}.

@@

The unified coeffect system uses a general notion of context shape, but it has been designed with 
structural and flat systems in mind. In this and the next section, we show how it captures 
the two coeffect systems from Chapter~\ref{ch:flat} and Chapter~\ref{ch:structural}.
